---
title: "Running GSVA in HPC and cloud environments"
author:
- name: Robert Castelo
  affiliation:
  - &idupf Dept. of Medicine and Life Sciences, Universitat Pompeu Fabra, Barcelona, Spain
  email: robert.castelo@upf.edu
abstract: >
  Here we illustrate how to use GSVA in an HPC environment, and how to access
  GSVA ranks stored in cloud object storage or on HTTPS servers, without the
  need to download them.
date: "`r BiocStyle::doc_date()`"
package: "`r pkg_ver('GSVA')`"
vignette: >
  %\VignetteIndexEntry{Running GSVA in HPC and cloud environments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteKeywords{GeneExpression, Microarray, RNAseq, GeneSetEnrichment, Pathway}
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
    fig_captions: yes
bibliography: GSVA.bib
---

**License**: `r packageDescription("GSVA")[["License"]]`

```{r setup, include=FALSE}
options(width=80)
knitr::opts_chunk$set(collapse=TRUE,
                      message=FALSE,
                      warning=FALSE,
                      comment="",
                      fig.align="center",
                      fig.wide=TRUE)
```

# Introduction

Splitting GSVA calculations in parallel on a multicore shared-memory computer is
directly supported in the functions `gsva()`, `gsvaRowNorm()`, `gsvaColRanks()`
and `gsvaColScores()`, by using the `BPPARAM` parameter with backends such
as `BiocParallel::MulticoreParam()` or `BiocParallel::SnowParam()`. However,
running GSVA in a high-performance computing (HPC) environment, distributing the
computational load across multiple independent nodes that do not share their
main memory, requires using some additional functions for that purpose. In this
vignette we illustrate this use case assuming our HPC environment uses the
[SLURM](https://slurm.schedmd.com) workload manager. Other workload managers
can be used, please consult the documentation of the `r Biocpkg("BiocParallel")`
package for more information, or open an
[issue](https://github.com/rcastelo/GSVA/issues/new/choose) in the GSVA GitHub
repo if you need help with a different workload manager.

In essence, we will be using the three-step GSVA pipeline, illustrated in the
scRNA-seq vignette, with the additional parameters `first` and `last`, which
allow the algorithm to run on a subset of rows (`gsvaRowNorm()`) or columns
(`gsvaColRanks()` and `gsvaColScores()`) of the input data, storing the
intermediate results using on-disk data structures, and merging them into a
final single output object that will store the enrichment scores using also
an on-disk data structure. We will not have to worry about how to split the
input data and merge the intermediate results. These two tasks will be executed
by the functions `gsvaMap()` and `gsvaReduce()`.

In the last section, we show how GSVA column ranks stored in Apache Parquet
format in cloud object storage, such as Amazon S3 or Google Cloud Storage
(GCS), or on HTTPS public servers, can be accessed lazily, downloading only the
parts of the data that each calculation needs.

# Using a SLURM workload manager

Let's assume we have run the first part of the scRNA-seq vignette, and we have
obtained a `SingleCellExperiment` object called `sce` with filtered and
normalized expression values, and a collection of gene sets stored in an object
called `gsets`.

We first build a parameter object using the function `gsvaParam()`, as we would
normally do.

```{r, eval=FALSE}
gsvapar <- gsvaParam(sce, gsets)
```
We would then build a `BatchtoolsParam` object, as described in the
corresponding
[vignette](https://bioconductor.org/packages/release/bioc/vignettes/BiocParallel/inst/doc/BiocParallel_BatchtoolsParam.html)
of the `r Biocpkg("BiocParallel")` package. The GSVA package, however, provides
a convenience function called `gsvaBatchtoolsSlurmParam()` that will build a
`BatchtoolsParam` object with some sensible settings for running GSVA in an HPC
environment with a SLURM workload manager.

```{r, eval=FALSE}
gsvabtpar <- gsvaBatchtoolsSlurmParam(partition="short", nodes=10,
                                      ncpus_per_task=4)
```
In this example, we are distributing the computational load across 10 nodes, where
in each of them calculations will be done in parallel using 4 CPU cores, sending
the corresponding jobs to a partition called `short`. By default, the function
`gsvaBatchtoolsSlurmParam()` has a first argument `dir="GSVAOUTPUT"`, which will
create a directory called `GSVAOUTPUT` in the current working directory, where
the intermediate results will be stored. The user can change this directory by
specifying a different value for the `dir` argument, but it is important that this
directory is accessible from all nodes in the HPC environment. The user is also
responsible for deleting the contents of this directory after the GSVA calculations
are finished. Each job saves its results under a temporary name ending in
`.partial`, which it renames once they are complete, so files or directories with
that ending left in this directory belong to jobs that did not finish, and can be
deleted once no job of those calculations is running.

The memory requested for each job, given by the `mem` argument of
`gsvaBatchtoolsSlurmParam()`, with a default value of `"10G"`, is also the maximum
main memory that GSVA will attempt to use in that job. Each job runs one R process
and, with more than one CPU core per job, as many parallel workers, and each of
these R processes takes about 0.8 GB by itself for loading GSVA and the packages it
depends on. Therefore, the memory requested for each job should be at least
`(1 + ncpus_per_task)` times 0.8 GB, i.e., 4 GB in the previous example, plus the
memory needed by the calculations on each chunk of the input data. When GSVA
estimates that its calculations require more memory than available, it gives a
warning before starting them, suggesting to use fewer CPU cores per job or to
request more memory.

The next step is to run the function `gsvaMap()`, which will submit the jobs to the
SLURM workload manager, and return a list object, where each of its elements will be
the path to the intermediate results of one chunk of the input data, saved in the
`GSVAOUTPUT` directory with the function `saveHDF5GSVA()`. Saving these results
to disk is the default behavior of `gsvaMap()` when jobs are sent to a workload
manager, such as SLURM, and can be changed with the `output` argument of
`gsvaMap()`.

```{r, eval=FALSE}
gsvamaprnorm <- gsvaMap(gsvaRowNorm, gsvapar, BTPARAM=gsvabtpar)
```
The final step is to merge the intermediate results into a single object, using
the function `gsvaReduce()`, which in this example will return a
`SingleCellExperiment` object with the row-normalized values stored in an
assay called `gsvarnorm`.

```{r, eval=FALSE}
gsvarnorm <- gsvaReduce(gsvamaprnorm)
```
To complete the three-step GSVA pipeline, we would then calculate the column
rank values using the function `gsvaColRanks()`, and finally the GSVA scores
using the function `gsvaColScores()`, as we would normally do in a single-node
environment, but using `gsvaMap()` and `gsvaReduce()`, this time for a more
compact coding in single line of code each.

```{r, eval=FALSE}
gsvaranks <- gsvaReduce(gsvaMap(gsvaColRanks, gsvarnorm, BTPARAM=gsvabtpar))
gsvaes <- gsvaReduce(gsvaMap(gsvaColScores, gsvaranks, BTPARAM=gsvabtpar))
```
Since the calculations of the column ranks and enrichment scores are both done
on the columns of the input data, we can reuse the output of `gsvaMap()`
obtained from the rank calculations, directly to the input of `gsvaMap()` for
the enrichment score calculations, without having to reduce the intermediate
results into a single object, avoiding the overhead of that unnecessary
reduction.

```{r, eval=FALSE}
gsvamapranks <- gsvaMap(gsvaColRanks, gsvarnorm, BTPARAM=gsvabtpar)
gsvaes <- gsvaReduce(gsvaMap(gsvaColScores, gsvamapranks, BTPARAM=gsvabtpar))
```
The current implementation using the `BatchtoolsParam` backend makes the
`gsvaMap()` function blocking, waiting for all jobs to finish before returning
a list object with the intermediate results. There are currently two ways to
circumvent this limitation. The first one is to use a terminal multiplexer such
as [tmux](https://github.com/tmux/tmux/wiki), log interactively into a node of
the HPC environment using a job submission command such as:

```{r, eval=FALSE}
srun --mpi=none --mem=10G --nodes=1 --ntasks-per-node=1 \
     --partition=short --pty bash -i
```
start an R session, call interactively the `gsvaMap()` function and detach the
tmux session to be able to log out and leave the jobs running in the background.
You may reattach to the tmux session later to check the progress of the jobs or
fetch the results. Note that the partition where the interactive job is
submitted must have a walltime limit that is long enough to allow the jobs to finish.

The second way is to submit a job with the R script that calls the `gsvaMap()`
function, which will submit the jobs to the SLURM workload manager.

## Resubmitting failed chunks

If the calculations on some of the chunks of the input data fail, `gsvaMap()`
gives a warning and returns a partial result, in which the elements of the
failed chunks are condition objects describing the errors. Such a partial
result cannot be given to `gsvaReduce()`, but it can be given to the `MAPREDO`
argument of a new call to `gsvaMap()`, with the same `FUN` and `inputData`
arguments, to compute only the failed chunks, using the same chunk boundaries
as in the previous call. This allows one, for instance, to resubmit the failed
chunks with more memory per task.

```{r, eval=FALSE}
gsvamapes <- gsvaMap(gsvaColScores, gsvamapranks, BTPARAM=gsvabtpar)
gsvabtpar2 <- gsvaBatchtoolsSlurmParam(partition="short", nodes=10,
                                       ncpus_per_task=4, mem="20G")
gsvamapes <- gsvaMap(gsvaColScores, gsvamapranks, BTPARAM=gsvabtpar2,
                     MAPREDO=gsvamapes)
gsvaes <- gsvaReduce(gsvamapes)
```
Since `gsvaMap()` saves the results of each chunk to disk as soon as it is
completed, chunks of jobs killed by the workload manager, for instance for
exceeding their wall time or memory limits, are also reported as failed in the
partial result, and can be resubmitted in the same way. If the R session that
called `gsvaMap()` ended before the jobs finished, for instance because it was
running in an interactive job that reached its own wall time limit, the
partial result is not available. In that case, we can give to the `MAPREDO`
argument the directory where `gsvaMap()` saved the results, which by default
is `GSVAOUTPUT`, and the chunks whose results are not found in that directory
will be resubmitted.

```{r, eval=FALSE}
gsvamapes <- gsvaMap(gsvaColScores, gsvamapranks, BTPARAM=gsvabtpar,
                     MAPREDO="GSVAOUTPUT")
```
To find the results of a previous call, `gsvaMap()` writes in that directory a
manifest file, whose name ends in `_manifest.rds`, and gives the results of
each chunk a file name that only depends on the input data, the output format
and the chunk boundaries. For this reason, calling `gsvaMap()` again with the
same input data and output format, without the `MAPREDO` argument, gives an
error when that directory already has the results of a previous call. The input
data is identified by its dimensions, dimension names and GSVA parameters, but
not by its values, so the user should not change the values of the input data
between the call to `gsvaMap()` and the one resubmitting the failed chunks.

Note that `gsvaMap()` may take a few minutes to return after the last job has
ended when some jobs were killed, because `r CRANpkg("batchtools")` only
considers that a job without results has ended after not finding it in the
queue of the workload manager in several consecutive checks. To shorten this
delay, `gsvaBatchtoolsSlurmParam()` writes in the directory given in its `dir`
argument a batchtools configuration file that makes those checks at most 30
seconds apart, instead of up to two minutes.

# Accessing remotely stored GSVA ranks

Column ranks calculated with `gsvaColRanks()` can be stored in a single file in
[Apache Parquet](https://parquet.apache.org) format with the function
`saveParquetGSVA()`. This requires having installed the CRAN package
[arrow](https://cran.r-project.org/package=arrow) [@richardson2026arrow].

```{r, eval=FALSE}
saveParquetGSVA(gsvaranks, "gsvaranks.parquet")
```
When the ranks are calculated with `gsvaMap()` in an HPC environment, as in the
previous section, the results of each worker can be directly saved in Parquet
format with the argument `output="Parquet"` in the call to `gsvaMap()`, and
then those results can be combined into a single file with `gsvaReduce()` and
stored in a single Parquet file with `saveParquetGSVA()`.

```{r, eval=FALSE}
gsvamapranks <- gsvaMap(gsvaColRanks, gsvarnorm, output="Parquet",
                        BTPARAM=gsvabtpar)
saveParquetGSVA(gsvaReduce(gsvamapranks), "gsvaranks.parquet")
```
The particular Parquet format file written by GSVA is organized to enable
efficiently reading blocks of columns, such as when we want to calculate
enrichment scores for our gene sets of interest. This is useful when the
Parquet file with GSVA ranks has been uploaded to a cloud storage bucket in
Amazon S3 or GCS, e.g., with the command-line tools `aws s3 cp` or
`gcloud storage cp`, because it can be lazily loaded with the function
`loadParquetGSVA()`, giving an `s3://` or `gs://` URI as the main file
argument, i.e., without downloading the whole Parquet file.

```{r, eval=FALSE}
gsvaranks <- loadParquetGSVA("s3://my-bucket/gsvaranks.parquet")
```
Loading the file only downloads its metadata, including the row and column
data of the original object, such as the `colData()` of a
`SingleCellExperiment`, while the ranks themselves are downloaded only when
they are needed, in the blocks of columns that each calculation reads. For
instance, we can calculate GSVA scores for a new collection of gene sets
without downloading the whole matrix of ranks at once, since they will be
read and processed by blocks of columns.

```{r, eval=FALSE}
gsvaes <- gsvaColScores(gsvaranks, geneSets=newgsets)
```
In the same way, calculating the enrichment data of a single gene set in a
single column, e.g., a single cell, with the function `gsvaEnrichment()`,
only downloads the chunk of columns that contains that column.

```{r, eval=FALSE}
res <- gsvaEnrichment(gsvaranks, column=10000, geneSet=newgsets[[1]])
```
Accessing the file requires an arrow installation with support for the
corresponding cloud storage, which can be checked with
`arrow::arrow_with_s3()` and `arrow::arrow_with_gcs()`^[Note that CRAN builds of arrow include S3 support but not GCS support. The latter can be obtained by installing arrow from R-universe; see https://arrow.apache.org/docs/r/articles/fs.html].
Files stored in private cloud storage buckets require the arrow package to
read the necessary credentials to access them. Please consult the corresponding
documentation for AWS and GCS on how to set up the credentials in your system.
Files in public S3 buckets can be accessed directly with the `s3://` URI, e.g.,
`s3://my-bucket/gsvaranks.parquet`, which already requests anonymous access,
while files in public GCS buckets should be specified with URIs of the form
`gs://anonymous@my-bucket/gsvaranks.parquet` to request anonymous access. See
`help(loadParquetGSVA)` for further details.

Parquet files with GSVA ranks deposited on public HTTPS servers can be also
lazily loaded in the same way, by giving their `https://` URL to
`loadParquetGSVA()`. In this case, the file is read using the API of the CRAN
package [duckdb](https://cran.r-project.org/package=duckdb)
[@muehleisen2026duckdb], which needs to be installed, and the HTTPS server must
support HTTP range requests, which most servers do.

```{r, eval=FALSE}
gsvaranks <- loadParquetGSVA("https://my-server.org/data/gsvaranks.parquet")
```
The first time a file is accessed through HTTPS, the duckdb package downloads
its extension `httpfs`, which GSVA keeps for subsequent R sessions. If the
certificate of the HTTPS server cannot be verified, e.g., because the server
does not send its intermediate certificates, a file with the certificates of
the certification authorities can be given with
`options(GSVA.ca_cert_file="<file>")`.

Since each block of columns is downloaded every time it is read, the time
to access the data depends mostly on the network, and it will be shorter
when the cloud storage bucket is located in a region close to where the
calculations run. If the same ranks are going to be processed many times, it
may be faster to download the file once and work with the local copy.

# Session information {.unnumbered}
Here is the output of `sessionInfo()` on the system on which this document was
compiled running pandoc `r rmarkdown::pandoc_version()`:

```{r session_info, cache=FALSE}
sessionInfo()
```

# References
